%********************************************************************
% Title: Excel for CLL
% Description: An Open Educational Resource for my CLL Excel class
% Author: George Self
% History: 
%   200320: Started This Puppy
%********************************************************************

\RequirePackage{fix-cm} % fix some latex issues see: http://texdoc.net/texmf-dist/doc/latex/base/fixltx2e.pdf
\documentclass[ twoside, openright, titlepage, numbers=noenddot, headinclude,
  %1headlines, % letterpaper a4paper, 
  footinclude=true, cleardoublepage=empty, abstractoff, 
  % <--- obsolete, remove (todo)
  BCOR=5mm, 
  %paper=a4,
  paper=letter, fontsize=11pt,
  % ngerman (New German) formats German umlauts (among other things)
  american
  %
]{scrreprt}

%*********************************************************************
% Note: Make all your adjustments in here
%*********************************************************************
\input{Excel-config}

%*********************************************************************
% Bibliographies
%*********************************************************************
%\addbibresource{Bibliography.bib}
%\addbibresource[label=ownpubs]{AMiede_Publications.bib}

%*********************************************************************
% Hyphenation
%*********************************************************************
%\hyphenation{put special hyphenation here}

%*********************************************************************
% The document
%*********************************************************************
\begin{document}
\frenchspacing
\raggedbottom
\selectlanguage{american} % american ngerman
%\renewcommand*{\bibname}{new name}
%\setbibpreamble{}
\pagenumbering{roman}
\pagestyle{plain}
%*********************************************************************
% Frontmatter
%*********************************************************************
%\include{FrontBackmatter/DirtyTitlepage}
\include{FrontBackmatter/Titlepage}
\include{FrontBackmatter/Titleback}
%\cleardoublepage\include{FrontBackmatter/Dedication}
\cleardoublepage\include{FrontBackmatter/Foreword}
%\cleardoublepage\include{FrontBackmatter/Abstract}
%\cleardoublepage\include{FrontBackmatter/Publications}
%\cleardoublepage\include{FrontBackmatter/Acknowledgments}
\pagestyle{scrheadings}
\cleardoublepage\include{FrontBackmatter/Contents}
%*********************************************************************
% Mainmatter
%*********************************************************************
\cleardoublepage\pagenumbering{arabic}
\ctparttext{This course is a basic introduction to Excel. The course is designed for students who have little or no Excel experience. This course starts with creating new workbooks and then covers entering, editing, and managing worksheets, including selecting cells, entering data, and using the Excel Ribbon. The course also covers creating and using mathematical formulas and adding Excel functions to worksheets. Students who complete this course will have an excellent grounding in Excel that they can use in their work, but they will also be prepared for more advanced topics.}
\part{Basic Excel}
\cleardoublepage
\include{Chapters/01Fundamental}
\include{Chapters/02Computations}
\include{Chapters/03Formulas}

\ctparttext{This course covers creating and using Excel charts to simplify data analysis. The course includes an emphasis on selecting an appropriate type of chart for the audience and data. The course next covers Excel tables and how they can be used to provide analysis tools like filters to simplify data. Finally, the course covers using multiple worksheets for complex data structures. Students who complete this course will be able to add graphic elements to worksheets and will be prepared for more advanced topics.}
\part{Intermediate Excel}
\cleardoublepage
\include{Chapters/04Charts} 
\include{Chapters/05Tables}
\include{Chapters/06MultipleSheets}
%
\ctparttext{This course covers importing and analyzing data using pivot tables and what-if analysis. It then covers using Excel to create predictions based on the data available. Finally, the course presents various ways that Excel can be used for statistical analysis of data. Students who complete this course will be prepared to use Excel for any data analysis function required in their work or studies.}
\part{Advanced Excel}
\cleardoublepage
\include{Chapters/07Data}
\include{Chapters/08Forecasting}
%\include{Chapters/09Statistics}

%*******************************************************************
% Sandbox
%*******************************************************************
%\include{Chapters/98Sandbox}
%\include{Chapters/97Template}

%*******************************************************************
% Backmatter
%*******************************************************************
%\appendix
%\renewcommand{\thechapter}{\alph{chapter}}
%\cleardoublepage
%\ctparttext{This part contains the Glossary, Bibliography, and Colophon.}
%\part{Appendix}
%\include{Chapters/00Appendix}
%*********************************************************************
% Other Stuff in the Back
%*********************************************************************
%\cleardoublepage\include{FrontBackmatter/Glossary}
%\cleardoublepage\include{FrontBackmatter/Bibliography}
%\cleardoublepage\include{FrontBackmatter/Declaration}
\cleardoublepage\include{FrontBackmatter/Colophon}
%*********************************************************************
% Game Over: Restore, Restart, or Quit?
%*********************************************************************
\end{document}
